### 2016-03-23 PaperThoughts.Rnw
more maths

\item \textbf{Permitted Changes} - First note a slight abuse of notation. The current state of the network, which was previously denoted $X(t)$ and $x$, will now be denoted $x^0$. Then, let $x$ instead represent the new, as yet undetermined network $X(t+1)$. So, given that a change, $x^0 \rightarrow x$ will occur, and that node $i$ has been selected to change a tie, there is a set of permitted changes, $\mathcal{A}_i(x^0) = \{\mathcal{A}(x^0) \cup \mathcal{A}_i^r(x^0)\}$, where $\mathcal{A}(x^0)$ is the current network's adjacency matrix, and $ \mathcal{A}_i^r(x^0) = \{x | x_{ij} = 1 - x_{ij}^0 \text{for only one } j \in \mathcal{N},  j \neq i, \}$.  So, the set of permitted changes includes the set of all adjacency matrices with all entries equivalent to the current adjacency matrix with the exception of one of the non-fixed, $n-1$ entries of the $i^{th}$ row and the current adjacency matrix, meaning that $|\mathcal{A}_i(x^0)| = n$. 

@@

Properties of intensity matrices dictate that the diagonal entry, where the state doesn't change, is defined as $q(x^y,x^y) = -\sum_{z\neq y} q(x^y, x^z)$ so that the row sum is 0. Thus, the rate of change between any two states that differ by only one tie from $i \to j$ is the product of the rate at which actor $i$ gets to change a tie and the probability that the tie that will change is the tie to node $j$.\footnote{Just to be clear, the change is from $x^y_{ij}$ to $x^z_{ij} = 1 - x^y_{ij}$.} Furthermore, the theory of continuous time Markov chains gives that the matrix of transition probabilities between observation times $t_{m-1}$ and $t_{m}$ is dependent only on the difference between timepoints, $t_m - t_{m-1}$ and is equal to $e^{(t_m - t_{m-1})Q}$, where $Q$ is the matrix defined above and $e^X$ for a real or complex square matrix $X$ is equal to $\sum_{k=0}^{\infty} \frac{1}{k!} X^k$. There is a package in \texttt{R} to do this, but I can't imagine doing this for any $n > 5$.  

@@

>>>>

\item \textbf{Permitted Changes} - First note a slight abuse of notation. The current state of the network, which was previously denoted $X(t)$ and $x$, will now be denoted $x^0$. Then, let $x$ instead represent the new, as yet undetermined network $X(t+1)$. So, given that a change, $x^0 \rightarrow x$ will occur, and that node $i$ has been selected to change a tie, there is a set of permitted changes, $\mathcal{A}_i(x^0) = \{\mathcal{A}(x^0) \cup \mathcal{A}_i^r(x^0)\}$, where $\mathcal{A}(x^0)$ is the current network's adjacency matrix, and $ \mathcal{A}_i^r(x^0) = \{x | x_{ij} = 1 - x_{ij}^0 \text{for only one } j \in \mathcal{N},  j \neq i, \}$.  So, the set of permitted changes includes the set of all adjacency matrices with all entries equivalent to the current adjacency matrix with the exception of one of the non-fixed, $n-1$ entries of the $i^{th}$ row and the current adjacency matrix, meaning that $|\mathcal{A}_i(x^0)| = n$. Additionally, define the complete set of possible changes from $x^0 \rightarrow x$ as $\mathcal{A}_{1:n}(x^0) = \{\mathcal{A}(x^0) \cup \mathcal{A}_1(x^0) \cup \mathcal{A}_2(x^0) \cup \cdots \cup \mathcal{A}_n(x^0)\}$. This is the set of all networks that differ from $x^0$ by only one tie and has size $|\mathcal{A}_{1:n}(x^0)| = n(n-1) + 1$. 

@@

Properties of intensity matrices dictate that the diagonal entry, where the state doesn't change, is defined as $q(x^y,x^y) = -\sum_{z\neq y} q(x^y, x^z)$ so that the row sum is 0. Thus, the rate of change between any two states that differ by only one tie from $i \to j$ is the product of the rate at which actor $i$ gets to change a tie and the probability that the tie that will change is the tie to node $j$.\footnote{Just to be clear, the change is from $x^y_{ij}$ to $x^z_{ij} = 1 - x^y_{ij}$.} Furthermore, the theory of continuous time Markov chains gives that the matrix of transition probabilities between observation times $t_{m-1}$ and $t_{m}$ is dependent only on the difference between timepoints, $t_m - t_{m-1}$ and is equal to $e^{(t_m - t_{m-1})Q}$, where $Q$ is the matrix defined above and $e^X$ for a real or complex square matrix $X$ is equal to $\sum_{k=0}^{\infty} \frac{1}{k!} X^k$. There is a package in \texttt{R} to do this matrix exponential operation, but I can't imagine doing this for networks for any $n > 5$.
\item \textbf{Moving from $x(t_{m-1})$ to $x(t_m)$} - First define $x(t_1)$ as the first network observed and $x(t_M)$ as the last network observed. Networks in between are written $x(t_{m})$ for $m = 2, \dots, M-1$. The observations $x(t_1), x(t_2), \dots, x(t_M)$ are embedded in the continuous-time Markov chain $\{X(t)|t_1 \leq t \leq t_M\}$. Moving from the observation $x(t_{m-1})$ to the observation $x(t_m)$ requires $C$ changes or steps in the Markov process $X(t)$ where $C = \sum_{i,j \in \mathcal{N}} |x_{ij}(t_m) - x_{ij}(t_{m-1})|$. Denote the $C-1$ timepoints in the the Markov process that correspond to the unobserved changes in the network as $T_1, T_2, \dots, T_{C-1}$ and define $T_0 = t_{m-1}$ and $T_C = t_{m}$. For each timepoint $T_c$, for $c = 1, \dots C-1$, there is a single actor, denoted $I_c$ that gets an opportunity to change at this timepoint. This actor changes one of its ties to one node, $J_c$. If there is no change, $I_c = J_c$. Formally, the pair $(I_c,J_c)$ is the only $(i,j)$ pair for which $x_{ij}(T_{c-1}) \neq x_{ij}(T_c)$ if such a pair exists, and $(I_c, J_c) = (I_c, I_c)$ otherwise. The triplet $(T_c, I_c, J_c)$ forms a stochastic process for $c = 1, \dots, C-1$ that, given the prior observation, $x(t_{m-1})$ for $m = 2, \dots, M$, completely determines $x(t)$ for $t_{m-1} < t < t_m$. 
\item \textbf{Augmenting Data} (see section on ML estimation) - The augmenting data consist of $C$ and the pairs $(I_c, J_c)$ for $c = 1, \dots, C-1$. (Note the timepoints $T_c$ are left out of the augmenting data.) The possible outcomes of the augmenting data are determined by the condition that if for a given tie, $x_{ij}(t_{m-1})$, the number of timepoints $1 \leq c \leq C-1$ at which $(i_c, j_c) = (i,j)$ is even if $x_{ij}(t_{m}) = x_{ij}(t_{m-1})$ and is odd if $x_{ij}(t_{m}) \neq x_{ij}(t_{m-1})$. 
\item \textbf{Sample Path} - The sample path is the stochastic process $V = ((I_c, J_c):c = 1, \dots, C-1)$. The elements of $V$ where $I_c = J_c$ are redundant. However, they are kept in the process because they help with computation of the likelihood. For notational convenience, define $x^{(c)} = x(T_C)$. Then, the networks $x^{(c)}$ and $x^{(c-1)}$ only differ in the element $(I_c, J_c) \in V$. 

@@

\section{Important Facts about SAOMs}
\begin{itemize}
\item In SAOMs, the observed networks in a longitudinal study, call them $x_{t_1}, \dots, x_{t_M}$, are discrete observations embedded in a continuous time process, $X(t)$. This means that the probability model is defined \emph{independently} of the observational design. Thus, the time between observations can be irregular and still be analyzed under this model.
\item The continuous-time Markov process $\{X(t) | t_1 \leq t \leq t_M\}$ has a distribution that consitutes an exponential family, which are of the form $f(y|\theta) = \exp\{\sum_{j=1}^s q_j(\theta) T_j(y) \} c(\theta) h(y)$ where $y$ is the data vector and $\theta$ is the scalar or vector parameter. This means the SAOMs can be viewed as ``incompletely observed exponential family" models. 
\item Inference is computer-intensive and time-consuming because of ``elaborate MCMC procedures" that must be implemented.
\item The definition of SAOMs means that data can be simulated directly from the probability distribution given the initial network state $x_{t_1}$. 
\item However, SAOMs are ``too complicated for the calculation of likelihoods or estimators in closed form." The distribution, however, can be easily simulated from, so that estimation can proceed by MCMC simulation.
\end{itemize}

\section{Maximum Likelihood Estimation}
Sketch of Algorithm:
\begin{enumerate}
\item Use the first network observation, $x(t_1)$, as a starting value. 
\item For $m = 2, 3, \dots, M$ where $M$ is the total number of observed networks, augment the observed data with random draws from the sequence of intermediate steps in the Markov process, $x(t)$, that could have led from $x(t_{m-1})$ to $x(t_m)$. i.e. draw from the set {\small $$\{(x(t^*_1), x(t^*_2), \dots x(t^*_{C-1}))' | x(t^*_1) \in \mathcal{A}_{1:n}(x(t_{m-1})) \& x(t^*_2) \in \mathcal{A}_{1:n}(x(t^*_1)) \& x(t^*_3) \in \mathcal{A}_{1:n}(x(t^*_2)) \& \cdots \& x(t_{m}) \in \mathcal{A}_{1:n}(x(t^*_{C-1})) \}$$ } where $C = \sum_{i,j \in \mathcal{N}} |x_{ij}(t_m) - x_{ij}(t_{m-1})|$. Note that $C$ in the number of single edge changes needed to get from $x(t_{m-1})$ to $x(t_m)$, so $C-1$ is the number of intermediate networks. These draws can be simulated using a Metropolis-Hastings algorithm. 
\item Use the draws from 2 as updates in a Robbins-Monro algorithm (?) to find the solution of the likelihood equation. 
\end{enumerate}
